---
title: "Getting started with `lidarMiner`"
author: 
- name: Hervé Guillon
date: "`r format(Sys.time(), '%d %B %Y')`" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Mining metadata from Rocky FTP

Most of the elevation data for the United States are hosted by the USGS on the [Rocky FTP](rockyftp.cr.usgs.gov:/vdelivery/Datasets/Staged/Elevation/1m/Projects).
Here, each project (and sometimes work packages within a project) is hosted in its own folder with three subfolders:

- `TIFF` which contains the actual tiles of the dataset as large `.tif` files;
- `browse` which contains `.jpg` previews for internet browsing platforms like [The National Map](https://viewer.nationalmap.gov/basic/);
- `metadata` which contains metadata for each tile as light-weight `.xml` files. 

These metadata files can be mined with the following  commands, here for California, as indicated by `CA`:

1. Calling `mkdir data` at the root of the repository will create the `/data` folder
1. `cd data` will change to `data/`
1. `lftp rockyftp.cr.usgs.gov:/vdelivery/Datasets/Staged/Elevation/1m/Projects` connects to Rocky FTP
1. `mirror --no-empty-dirs -n . . -i "(_CA_).*(xml)$"` downloads the files matching the regular expression `"(_CA_).*(xml)$"`.
	+ You can kill this command once it has run past the state of interest.
	+ It takes around 20 min to download all the metadata for California.

# Setup

For the rest of this example, we will use the packages: `lidarMiner` and `raster` (which loads `sp`).

```{r setup}
library(lidarMiner)
library(raster)
```

# Executing `lidarMiner` functions

This chunk executes all functions from `lidarMiner`, listing the files at a specified path, reading the `.xml` files and extracting the spatial information into a `SpatialPolygonsDataFrame`.
The files required files are included with the package and stored in `/inst/extdata` that we access with `system.file`.

```{r executing lidarMiner functions, fig.width = 6, fig.height = 6}
list_files <- list_lidar_files(system.file("extdata", package = "lidarMiner"))
lidar_df <- get_lidar_df(list_files)
lidar_bbox <- get_lidar_bbox(lidar_df)
plot(lidar_bbox)
```

# Leveraging the bounding box of the lidar datasets to select the target tiles

First, let's load a `region_boundary` shapefile and a `sampling_locations` shapefile.
The first describes the boundary of an area of interest, here the North Central Coast of California, USA.
The second holds the locations of points of interest within the region, here field location of channel morphology sampling.


```{r loading test data, fig.width = 6, fig.height = 6}
region_boundary <- shapefile(system.file("extdata", "region_boundary.shp", package = "lidarMiner"))
sampling_locations <- shapefile(system.file("extdata", "sampling_locations.shp", package = "lidarMiner"))
plot(region_boundary)
plot(sampling_locations, add = TRUE)
```

It might be useful to add some buffering around the point if you know how you will use the lidar datasets.
For example, as I know that I will search for topographical information in squared tile $500\times500$-m, I add a circular buffer of $\sqrt{2}\cdot 250$-m around each location.
This ensures that I capture enough lidar data to proceed with my downstream analysis. 

```{r adding some buffer}
sampling_locations_polygons <- buffer(sampling_locations, width = sqrt(2) * 250, dissolve = FALSE)
```

Because the `sampling_locations` and `region_boundary` do not share the same CRS as `lidar_bbox`, this chunk handles some CRS conversions with `spTransform`.

```{r handling crs conversion}
region_boundary <- spTransform(region_boundary, crs(lidar_bbox))
sampling_locations <- spTransform(sampling_locations, crs(lidar_bbox))
sampling_locations_polygons <- spTransform(sampling_locations_polygons, crs(lidar_bbox))
```

We are now ready to perform a spatial intersect.
We perform a spatial intersect with the region boundary to speed up the subsequent intersect with the sample locations.


```{r spatial intersect, warning = FALSE, fig.width = 6, fig.height = 6}
lidar_bbox <- intersect(lidar_bbox, region_boundary)
lidar_sampling_locations <- intersect(lidar_bbox, sampling_locations_polygons)
unique_networkr <- as.character(unique(lidar_sampling_locations$networkr))
plot(lidar_bbox)
plot(lidar_bbox[lidar_bbox$networkr %in% unique_networkr, ], col = "green", add = TRUE)
plot(sampling_locations_polygons, add = TRUE)
```

You can note that in the center of the region, two tiles have been selected thanks to the buffering.

Finally, the `networkr` field of the resulting `lidar_sampling_locations` holds the URLs of the tiles we need.
```{r displaying urls}
head(unique_networkr)
```